---
title: "Week 6: Organizing your R code and data into a package"
output:
  html_document:
    toc: true
    include:
      after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(1234)
```

```{r chunk1, echo=FALSE, message=FALSE, warning=FALSE}
library(kableExtra)
dt <- data.frame("Compartmentalized", "Documented", "Extendible", "Reproducible", "Robust")
kable(dt, col.names=NULL) %>%
  kable_styling(full_width = TRUE) %>%
  row_spec(1, bold = FALSE, color = "white", background = "blue") %>%
  column_spec(column = 1:5, width = "20%")
```


Part 3, Sharing your package.


## Sharing your package on GitHub

Once you have your package on GitHub, it is easy for people to install using

```
library(devtools)
install_github("youraccount/MyNewPackage")
```

For example to install the package on 'RVerse-Tutorials', you would use
```
install_github("RVerse-Tutorials/TestPackage")
```
Note, {remotes} is another option that you'll see instead of {devtools}.

### Making sure people use the latest version

You can create releases on GitHub and give people these installation instructions:

```
devtools::install_github("youraccount/MyNewPackage@*release")
```

## Examples of real data packages

* [SardineForecast](https://github.com/eeholmes/SardineForecast) A complex package with lots of different types of data.
* [VRData](https://nwfsc-math-bio.github.io/VRData/) This is a data repo that I am developing for NWFSC application.

## How do I get my package on GitHub?

[PSAW II Workshop Introduction to GitHub](https://rverse-tutorials.github.io/PSAW-2022/)

## Data package collaboration

In a team application, you'll be dividing up the work.

* **The package development**. R coder and knows how to design a package.
* **Data**. Team members who add files to `data-raw` in whatever format your team uses.
* **Documentation**. Multiple team members might be involved in editing the files in the `R` directory. Note, in **VRData** the documentation headers are in `data-raw` and I have code that processes that into the files in `R`.
* **Package maintainer**. Not necessarily the package developer. This person makes sure releases are posted to GitHub.




    
    




