---
title: "Introduction to RStudio Cloud for Collaborating"
output:
  html_document:
    toc: true
    include:
      after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(kableExtra)
dt <- data.frame("Compartmentalized", "Documented", "Extendible", "Reproducible", "Robust")
kable(dt, col.names=NULL) %>%
  kable_styling(full_width = TRUE) %>%
  row_spec(1, bold = FALSE, color = "white", background = "blue") %>%
  column_spec(column = 1:5, width = "20%")
```

# Overview

In this half of the workshop, I'll go over some tips for how you can create workspaces for collaborating or for teaching.

* Create a project that students open from a link
* Create a shared workspace where student create projects
* Special software: Stan, TMB, etc

## Create a project for students to use

In this case, we create a project that students will open from a link.

https://github.com/RVerse-Tutorials/TestPackage

Here is an example. Click this and you'll open a project that you can work with. https://rstudio.cloud/project/3430079

Let's see how I created that.

### Pros

* Simple
* Doesn't burn up your project hours
* Doesn't matter how many students you have
* They can work in a temporary copy or save to their account

### Cons

* You can't see their work
* Not for collaborating, rather for sharing
* If you have a Premium account, they will be much more limited in RAM

## Create a workspace for students to use

Demo

### Pros

* If you have money or institutional support, you can have lots of RAM and CPU and students can share this more powerful workspace.
* You can see their work and give feedback or problem-solve. But you can't both be in the space at the same time.
* Good way to collaborate with a small group on the same project (but not at the same time).

### Cons

* On the free account, it works up to 4 members. You need to pay a not insignificant amount of money to get a lot of students.
* Not so sure it would scale up to 100s of students unless you are at a university (which gets steep discounts). Your RAM limit would get hit.

## Installing special software

One of the great features of using a cloud platform is that you can install software that would be a major hassle for your students to install:

* You have no idea what operating system they are using or its vintage
* They might not have admin access
* Install might be (substantially) different on different operating systems
* There could be many steps involved
* They might be moving between machines (home, work, computer lab)
* You might want to demo on a machine without having to install anything

Sadly, you are very likely to run into the limitation of 1 gig of RAM in the Free and Free-Plus tiers. The next tier is Premium and that is $99 a month. Better option is using AWS and setting up an RStudio free server.

### Examples

https://rstudio.cloud/learn/guide/system-packages

* JAGS works because it is already installed: https://rstudio.cloud/project/2228346

* sdmTMB throws an error due to the 1 gig RAM on free: https://rstudio.cloud/project/3429862

* Stan also runs in the 1 Gig RAM problem: 
    - https://rstudio.cloud/project/2228084
    - Using Kaggle is an option that has been suggested for short-term demos: https://www.kaggle.com/eeholmes/rstan  Watchg the RAM and you'll see that it is going over 1 Gig.


## Cost

For occasional use and code-sharing, the free plan should be fine. At most, you might need the Plus plan ($5 per month). https://rstudio.cloud/plans/free

